---
title: "Homework 1"
author: "Julio Cesar Enciso-Alva"
date: "January XX, 2019"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require('magrittr')
require('dplyr')
require('data.table')
```

**Note:** This documents uses the package `magrittr`, `dplyr` and `data.table`. It also uses the files `DistMat.csv`, distance matrix from question 3, and `T11-9.txt`, local copy of `cereal` database.

## Question 1

### a)

The multiplication will be calculated using the matrix product operator `%*%`.
In order to do that, the given matrices are saved to the variables `A` and `B`.

```{r}
A = matrix(c(7,9,20, 2,4,10),ncol=3)
B = matrix(c(1,7,12,40, 2,8,13,20, 3,9,14,21),ncol=4)
```

Then, the multiplication is calculated and displayed.

```{r}
A.B = A %*% B
A.B
```

### b)

The matrix $G$ will be constructed using the spectral decomposition of $A$; if we have an ortonormal matrix $C$ and a diagonal matrix $D$ such that 

\[
A = C D C'
\]

then its *square root* can be constructed (and will be proposed as the solution) using the following expression
\[
G = C D^\frac{1}{2} C'
\]

As the first step for doing so, the matrix $A$ is saved to the variable `A`.

```{r}
A = matrix(c(2,1,0, 1,2,0, 0,0,2),ncol=3)
```

The spectral decomposition of $A$ is found using the function `base::eigen`. It is important to recall that the matrices $C$ and $D$ such that $A = C D C'$ can be constructed as

\begin{align*}
C &= [x_1,x_2,x_3] \\
D &= \textrm{diag}(\lambda_1,\lambda_2,\lambda_3)
\end{align*}

where $x_i$ are the eigenvectors of $A$, and $\lambda_i$ are their corresponding eigenvalues.
Both values are saved to the variable `SD`.

```{r}
SD = eigen(A)
C = SD$vectors
D = diag(SD$values)
```

The matrix G, the **proposed solution**, is constructed as described before:

```{r}
G = C %*% sqrt(D) %*% t(C)
G
```

In order to verify that $G^2=A$, both $G^2$ and $A$ are displayed

```{r}
A
```

```{r}
G %*% G
```

### c)

The expression requested is
\[
\Phi(x) = \int_{-\infty}^x \frac{1}{\sqrt{2\pi}} e^{-\frac{z^2}{2}}\, dz
\]

## Question 2

To perform the clusterings, the given distance matrix was saved on the file `DistMat.csv`; the data is then loaded and saved to the variable `d`.
For ease of use, only the lower half of the matrix is in the file.

```{r}
d = read.csv('./DistMat.csv',header=F) %>%
  as.matrix() %>% as.dist()
```

The dendrograms are generated using the function `stats::hclust`, changing the method of clustering. 

```{r out.height='80%',echo=FALSE}
d %>% stats::hclust(method='single') %>% plot(sub=NA,main='Dendrogram using Single linking')
```

```{r out.height='80%',echo=FALSE}
d %>% stats::hclust(method='complete') %>% plot(sub=NA,main='Dendrogram using Complete linking')
```

```{r out.height='80%',echo=FALSE}
d %>% stats::hclust(method='average') %>% plot(sub=NA,main='Dendrogram with Average linking')
```

The three dendrograms have the same structure, but the nodes have different heights. This indicates that the clusters are robust.

## Question 3

The dataset `cereal` is loaded from a local copy using the suggested code; the column names are also read from a file. The variables `Manufacturer` and `Group are dropped`.

```{r include=FALSE}
cereal = fread("./T11-9.txt", data.table = F)
colnames(cereal) = scan(what="character", n=11, file='ColumnNames.txt')

rownames(cereal) = cereal[,1]
cereal = cereal[,-1]

cereal = colnames(cereal) %>% setdiff(c('Manufacturer','Group')) %>% cereal[.]
```

### a)

The distance between vectors is calculated using the function `stats::dist` and saved to the variable `d`.

```{r}
d = stats::dist(cereal, method='euclidean')
```

### b)

The clustering is achieved using the funcion `stats::hclust`, using different methods.

```{r out.height='350%', echo=FALSE}
d %>% stats::hclust(method='single') %>% plot(sub=NA,main='Dendrogram using Single linking')
```

```{r out.height='350%',echo=FALSE}
d %>% stats::hclust(method='centroid') %>% plot(sub=NA,main='Dendrogram using Centroid linking')
```

The clusters differ with the two methods. There are inversions on the *centroid* clustering, possible related to ~~healthy~~ low-sugar cereals. This difference can be explained by the effect of the variables `Calories`, `Sodium` and `Potassium`; those variables have very high values because of the scale.

## Question 4

The dataset `cereal` is loaded from a local copy using the suggested code; the column names are also read from a file. The variables `Manufacturer` and `Group` are dropped. **Note:** This process was done in *Question 3*, but is done again to keep the code independent.

```{r include=FALSE}
cereal = fread("./T11-9.txt", data.table = F)
colnames(cereal) = scan(what="character", n=11, file='ColumnNames.txt')

rownames(cereal) = cereal[,1]
cereal = cereal[,-1]

cereal = colnames(cereal) %>% setdiff(c('Manufacturer','Group')) %>% cereal[.]
```

The K-means clustering is done using the function `stats::kmeans`, and saved to a table.

```{r}
set.seed(1)
K2 = kmeans(cereal,centers=2)$cluster
K3 = kmeans(cereal,centers=3)$cluster
```

## Question 5

The required vector is generated and saved to the variable `x`, and then the required tasks are completed.

```{r, include=FALSE}
set.seed(1) 
x = rnorm(10000)
```

### a)

There are
`r length( x[ -2<x & x<2 ] )`
values in `x` between -2 and 2.
This value was calculated using the following code

```{r,eval=FALSE}
length( x[ -2<x & x<2 ] )
```

### b)

The indexes of `x`, whose values lie are on the set $(-\infty,-3]\cup[3,\infty)$, are saved to the variable `IND` and displayed.

```{r}
IND = c( which(x<=-3 | x>=3) )
IND
```

### c)

The histogram is constructed using the function `hist`.

```{r,echo=FALSE}
hist(x)
abline(v=median(x),lty=2,lwd=3,col='red')
```

## Question 6

The dataset `cereal` is loaded from a local copy using the suggested code; the column names are also read from a file. The variable `Group` is dropped. 

**Note:** This process was done in *Question 3* and *Question 4*, but is done again to keep the code independent, and also because variable `Manufacturer` is not dropped in this question.

```{r include=FALSE}
cereal = fread("./T11-9.txt", data.table = F)
colnames(cereal) = scan(what="character", n=11, file='ColumnNames.txt')

rownames(cereal) = cereal[,1]
cereal = cereal[,-1]

cereal = colnames(cereal) %>% setdiff(c('Group')) %>% cereal[.]
```

### a)

The table of means is constructed using the following code, and the displayed.

```{r}
cereal %>%
  dplyr::filter( Protein>=3 & Sugar<=9 ) %>%
  dplyr::group_by(Manufacturer) %>%
  dplyr::summarize(Mean_Calories=mean(Calories)) %>%
  knitr::kable()
```

### b)

The variable `Healthy` is constructed using the following code:

```{r}
cereal$Healthy = 1*(cereal$Protein>=3 & cereal$Sugar<=9)
```